Type theory in type theory with single substitutions

### 2024-09-16 types_2024_postproc/p.tex
stuff

\newcommand{\ra}{\rightarrow}
\newcommand{\Ra}{\Rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\ext}{\mathop{\triangleright}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\cd}{\mathsf{c}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}} %from the book
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\fst}{\mathsf{fst}}
\newcommand{\snd}{\mathsf{snd}}
\newcommand{\Lift}{\mathsf{Lift}}
\newcommand{\mk}{\mathsf{mk}}
\newcommand{\un}{\mathsf{un}}

@@

>>>>

\newcommand{\ra}{\rightarrow}
\newcommand{\Ra}{\Rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\Tel}{\mathsf{Tel}}
\newcommand{\Tms}{\mathsf{Tms}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\ext}{\mathop{\triangleright}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\cd}{\mathsf{c}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}} %from the book
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\fst}{\mathsf{fst}}
\newcommand{\snd}{\mathsf{snd}}
\newcommand{\Lift}{\mathsf{Lift}}
\newcommand{\mk}{\mathsf{mk}}
\newcommand{\un}{\mathsf{un}}

@@

\section{stuff}

There are equations in B-systems that cannot be derived from the equations of
SSC. The following four equations are such, they are versions of equations in
section \ref{sec:ssc} lifted under arbitrary amount of $\blank^+$s.
\begin{equation}\label{eq:lifted}
\begin{aligned}
  B[\p^{+^n}][(\gamma^+)^{+^n}] & = B[\gamma^{+^n}][\p^{+^n}] \\
  B[\p^{+^n}][\langle a \rangle^{+^n}] & = B \\
  B[\langle a \rangle^{+^n}][\gamma^{+^n}] & = B[(\gamma^+)^{+^n}][\langle a[\gamma] \rangle^{+^n}] \\
  B[(\p^+)^{+^n}][\langle\q\rangle^{+^n}] & = B
\end{aligned}
\end{equation}
In addition, we need the corresponding lifted equations for terms. Note that we
do not have the non-lifted term versions of the last two equations in SSC.

To formally state the equations in (\ref{eq:lifted}), we define telescopes as an
inductive type together with a recursive operation to append telescopes to
contexts.
\begin{alignat*}{10}
  & \Tel && : \Con\ra\Set \\
  & \blank+\blank && : (\Gamma:\Con)\ra\Tel\,\Gamma\ra\Con \\
  & \diamond && : \Tel\,\Gamma \\
  & \blank\ext\blank && : (\Omega:\Tel\,\Gamma)\ra\Ty\,(\Gamma+\Omega)\ra\Tel\,\Omega
\end{alignat*}
The appending operation is defined recursively on telescopes.
\begin{alignat*}{10}
  & \Gamma+\diamond && := \Gamma \\
  & \Gamma+(\Omega\ext A) && := (\Gamma+\Omega)\ext A
\end{alignat*}
We then define a lifting operation over any telescope mutually with an
instantiation operation on telescopes by recursion on telescopes.
\begin{alignat*}{10}
  & \blank[\blank] && : \Tel\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Tel\,\Delta \\
  & \blank^{+^\Omega} && (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta+\Omega[\gamma])\,(\Gamma+\Omega)
\end{alignat*}

If we try to prove the equations in (\ref{eq:lifted}) by induction on the types
and terms, we run into difficulties in the instantiation case, as we would have
to commute the instantiations without the equations to do so. Therefore, we
first $\alpha$-normalize the syntax to compute away all the instantiations, then
prove the equations by induction on the $\alpha$-normal forms.

We define $\alpha$-normal forms as inductive predicates in Prop

separate weakening and single substitution

Define instantiation operations by recursion on $\alpha$-normal forms

Instead of doing induction on the syntax for each of the equations in
(\ref{eq:lifted}), we define a general lemma which can lift any equation between
instantiations over a telescope.

We define $\Sub^*$ to be $\Sub$ with freely added identity and composition
operations. All instantiation operations, lifting operations, and substitution
rules can be redefined for $\Sub^*$.

\begin{lemma}
  Given and $\sigma, \delta : \Sub^*\,\Delta\,\Gamma$, if $A[\sigma] =
  A[\delta]$ and $x[\sigma] = x[\delta]$ for any $A:\Ty\,\Gamma$ and variable
  $x$ in $A$, then
  \begin{itemize}
    \item $\Omega[\sigma] = \Omega[\delta]$ for $\Omega:\Tel\,\Gamma$
    \item $A[\sigma^{+^\Omega}] = A[\delta^{+^\Omega}]$ for $A:\Ty\,(\Gamma+\Omega)$
    \item $a[\sigma^{+^\Omega}] = a[\delta^{+^\Omega}]$ for $a:\Tm\,(\Gamma+\Omega)\,A$
  \end{itemize}
\end{lemma}
\begin{proof}
  asd
\end{proof}

We can simulate parallel substitutions using $\Sub^*$ as iterated single
substitutions, however it does not satisfy the equations of parallel
substitutions. Thus we define parallel substitutions as a list of terms with a
map into $\Sub^*$, and reuse instantiation operation of $\Sub^*$ to convert it
into a sequence of instantiations of single substitutions.
\begin{alignat*}{10}
  & \Tms && : \Con\ra\Con\ra\Con \\
  & \llcorner\blank\lrcorner && : \Tms\,\Delta\,\Gamma \to \Sub^*\,\Delta\,\Gamma \\
  & \upvarepsilon && : \Tms\,\Gamma\,\diamond \\
  & \blank,\blank && : (\gamma:\Tms\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\llcorner\gamma\lrcorner])\ra\Tms\,\Delta\,(\Gamma\ext A) \\
  & \llcorner\upvarepsilon\lrcorner && := id\circ\dots\circ\p \\
  & \llcorner\gamma,a\lrcorner && := \llcorner\gamma\lrcorner^+\circ\langle a \rangle
\end{alignat*}
